---
title: "STA 325 Final Project Code"
author: "Calleigh Smith, Hannah Bogomilsky, Hugh Esterson, Maria Henriquez, Mariana Izon"
date: "November 23, 2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-packages, message=FALSE}
library(readr)
library(dplyr)
library(tidyverse)
library(gridExtra)
library(mgcv)
library(patchwork)
library(MASS)
library(EnvStats)
library(tree)
library(randomForest)
library(gbm)
library(bst)
library(plyr)
library(caret)
```

# Data Load-In and Initial Cleaning

```{r load-clean-data, warning=FALSE, message=FALSE}
# read data
flights <- read_csv("data/flights.csv")

# find unique airlines, destinations, and types of delays
unique(flights$OP_CARRIER)
unique(flights$DEST)

# mutate delays and filter out NA arrival delays
flights <- flights %>% 
  mutate(CARRIER_DELAY = case_when(CARRIER_DELAY > 0 ~ 1,
                                   TRUE ~ 0),
         WEATHER_DELAY = case_when(WEATHER_DELAY > 0 ~ 1,
            TRUE ~ 0),
         NAS_DELAY = case_when(NAS_DELAY > 0 ~ 1,
                               TRUE ~ 0),
         SECURITY_DELAY = case_when(SECURITY_DELAY > 0 ~ 1,
                                    TRUE ~ 0),
         LATE_AIRCRAFT_DELAY = case_when(LATE_AIRCRAFT_DELAY > 0 ~ 1,
                                         TRUE ~ 0)) %>% 
  filter(!is.na(ARR_DELAY))
```

```{r data-glimpse}
# glimpse data
flights
```

\newpage

# Exploratory Data Analysis
## Individual Predictor Variable EDA

### Taxi Histograms

```{r taxi-plots, warning=FALSE}
# plot untransformed predictor taxi_in
pTAXI_IN <- ggplot(data = flights, aes(x = TAXI_IN)) +
  geom_histogram(binwidth = 5, fill = "#FFFF00", color = "#002D72", alpha = .7) +
  labs(x = "Time to Taxi In",
       y = "Frequency",
       title = "Histogram of TAXI_IN") + 
  theme(plot.title = element_text(size = 10,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))
  
# plot untransformed predictor taxi_out
pTAXI_OUT <- ggplot(data = flights, aes(x = TAXI_OUT)) +
  geom_histogram(binwidth = 5, fill = "#FFFF00", color = "#002D72", alpha = .7) +
  labs(x = "Time to Taxi Out",
       y = "Frequency",
       title = "Histogram of TAXI_OUT") + 
  theme(plot.title = element_text(size = 10,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))
  
# log transform taxi_in and taxi_out
flights$log_TAXI_OUT <- log(flights$TAXI_OUT)
flights$log_TAXI_IN <- log(flights$TAXI_IN)

# plot log transformed taxi_out
plog_TAXI_OUT <- ggplot(data = flights, aes(x = log_TAXI_OUT)) +
  geom_histogram(binwidth = .25, fill = "#FFFF00", color = "#002D72", alpha = .7) +
  labs(x = "Log-Time to Taxi Out",
       y = "Frequency",
       title = "Histogram of log(TAXI_OUT)") + 
  theme(plot.title = element_text(size = 10,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))

# plot log transform taxi_in
plog_TAXI_IN <- ggplot(data = flights, aes(x = log_TAXI_IN)) +
  geom_histogram(binwidth = .25, fill = "#FFFF00", color = "#002D72", alpha = .7) +
  labs(x = "Log-Time to Taxi Out",
       y = "Frequency",
       title = "Histogram of Log(TAXI_IN)") + 
  theme(plot.title = element_text(size = 10,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))

pTAXI_IN + plog_TAXI_IN
pTAXI_OUT + plog_TAXI_OUT
```

### Days of Month and Week

```{r date-plots}
# plot predictor DAYS_OF_MONTH
pDOM <- ggplot(data = flights, aes(x = DAY_OF_MONTH)) +
  geom_histogram(binwidth = 1, fill = "#E81828", color = "#002D72", alpha = .8) +
  labs(x = "Days of Month",
       y = "Frequency",
       title = "Histogram of Days of Month") +
    theme(plot.title = element_text(size = 10,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))

# plot predictor DAY_OF_WEEK
pDOW <- ggplot(data = flights, aes(x = DAY_OF_WEEK)) +
  geom_histogram(binwidth = 1, fill = "#E81828", color = "#002D72", alpha = .8) +
  labs(x = "Day of Week",
       y = "Frequency",
       title = "Histogram of Days of Week") +
    theme(plot.title = element_text(size = 10,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))

pDOM + pDOW
```

### Destination Locations

Origin is all JFK, but we could consider the different destination locations. 

```{r DEST-plot}
# plot destinations in CA
pDEST <- ggplot(data = flights, aes(x = DEST)) +
  geom_bar(fill = "#40E0D0", color = "#002D72", alpha = .7) +
  labs(x = "Destination",
       title = "Bar Plot of Destinations") + 
  theme(plot.title = element_text(size = 12,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))

pDEST
```

### Airlines

```{r airline-plot}
# plot airline carriers
pLINE <- ggplot(data = flights, aes(x = OP_CARRIER)) +
  geom_bar(fill = "#40E0D0", color = "#002D72", alpha = .7) +
  labs(x = "Airline",
       title = "Bar Plot of Airlines") + 
  theme(plot.title = element_text(size = 12,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))

pLINE

# plot airlines by destination
pLINEDEST <- ggplot(data = flights, aes(x = DEST, fill = OP_CARRIER)) +
  geom_bar() + 
  labs(x = "Destination",
       title = "Bar Plot of Airlines by Destination",
       fill = "Airline")
    theme(plot.title = element_text(size = 12,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))
    
pLINEDEST
```

### Depart Delay Histogram

```{r departdelay-plot, warning = FALSE}
# plot DEP_DELAY
pDEPDELAY <- ggplot(data = flights, aes(x = DEP_DELAY)) +
  geom_histogram(binwidth = 4, fill = "#e9c2ed", color = "#002D72", alpha = 0.7) +
  xlim(-25, 50) +
  labs(x = "Departure Delay",
       y = "Frequency",
       title = "Histogram of DEP_DELAY") +
  theme(plot.title = element_text(size = 12,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))

pDEPDELAY
```

```{r delaytype-plot}
# plot types of delays
p1 <- ggplot(data = flights, aes(x = CARRIER_DELAY)) +
  geom_bar(fill = "#E81828", color = "#002D72") +
  labs(title = "Carrier Delay")

#p2 <- ggplot(data = flights, aes(x = WEATHER_DELAY)) +
  #geom_bar(fill = "#E81828", color = "#002D72") +
  #labs(title = "Weather Delay")

p3 <- ggplot(data = flights, aes(x = NAS_DELAY)) +
  geom_bar(fill = "#E81828", color = "#002D72") +
  labs(title = "NAS Delay")

p4 <- ggplot(data = flights, aes(x = SECURITY_DELAY)) +
  geom_bar(fill = "#E81828", color = "#002D72") +
  labs(title = "Security Delay")

p5 <- ggplot(data = flights, aes(x = LATE_AIRCRAFT_DELAY)) +
  geom_bar(fill = "#E81828", color = "#002D72") +
  labs(title = "Late Aircraft Delay")

grid.arrange(p1,p3,p4,p5, nrow = 3)
```

From this EDA of the categorical variables, we probably should not perform analysis with `SECURITY_DELAY` since all of them are classified as 0. 

Furthermore, only 9 flights are classified with a weather delay, so it may not be good for our model to include this as a variable for right now. 

Overall, the categorical delay predictors I would think we could use are: Carrier Delay, NAS Delay, and Late Aircraft Delay

## Response Variable EDA


```{r ARRDELAY-plot, warning = FALSE}
# plot ARR_DELAY
pARRDELAY <- ggplot(data = flights, aes(x = ARR_DELAY)) +
  geom_histogram(binwidth = 20, fill = "#002D72", color = "#E81828", alpha = 0.8) +
  xlim(-100, 150) +
  labs(x = "Arrival Delay",
       y = "Frequency",
       title = "Histogram of ARR_DELAY") + 
  theme(plot.title = element_text(size = 12,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))

pARRDELAY
```


## Predictors vs. Response EDA

### TAXI_IN / TAXI_OUT vs. ARR_DELAY
```{r TAXI-resp}
p6 <- ggplot(data = flights, aes(y = ARR_DELAY, x = TAXI_IN)) +
  geom_point() +
  labs(title = "Arrival Delay vs Taxi_In")

p7 <- ggplot(data = flights, aes(y = ARR_DELAY, x = TAXI_OUT)) +
  geom_point() +
  labs(title = "Arrival Delay vs Taxi_Out")



plog6 <- ggplot(data = flights, aes(y = ARR_DELAY, x = log_TAXI_IN)) +
  geom_point() +
  labs(title = "Arrival Delay vs log(Taxi_In)")

plog7 <- ggplot(data = flights, aes(y = ARR_DELAY, x = log_TAXI_OUT)) +
  geom_point() +
  labs(title = "Arrival Delay vs log(Taxi_Out)")

p6 + plog6
p7 + plog7
```

These plots above suggest that we may want to transform the variables at some point. 

### DEST vs. ARR_DELAY

```{r DEST-resp}
ggplot(data = flights, aes(y = ARR_DELAY, x = DEST)) +
  geom_boxplot() +
  labs(x = "Destination",
       title = "Arrival Delay vs Destination")
```

### DAY_OF_WEEK vs. ARR_DELAY

```{r DOW-resp}
p8 <- ggplot(data = flights, aes(y = ARR_DELAY, x = DAY_OF_WEEK)) +
  geom_point() +
  labs(title = "Arrival Delay vs Day of Week")

p9 <- ggplot(data = flights, aes(y = ARR_DELAY, group = DAY_OF_WEEK)) +
  geom_boxplot() +
  labs(title = "Arrival Delay vs Day of Week")

p8 + p9
```

### ARR_DELAY and DAY_OF_MONTH

```{r DOM-resp}
p10 <- ggplot(data = flights, aes(y = ARR_DELAY, x = DAY_OF_MONTH)) +
  geom_point() +
  labs(title = "Arrival Delay vs Day of Month")

p11 <- ggplot(data = flights, aes(y = ARR_DELAY, group = DAY_OF_MONTH)) +
  geom_boxplot() +
  labs(title = "Arrival Delay vs Day of Month")

p10 + p11
```

\newpage

# Additional Data Cleaning

## New Bounds and Variable
```{r data-cleaned}
# take only SFO/LAX since all 4 carriers fly there
# create TYPE_DELAY as a factor variable for type of delays
flights <- flights %>% 
  filter(DEST == "SFO" | DEST == "LAX") %>% 
  mutate(TYPE_DELAY = case_when(NAS_DELAY == 1 ~ "NAS",
                                CARRIER_DELAY == 1 ~ "CARRIER",
                                LATE_AIRCRAFT_DELAY == 1 ~ "LATE_AIRCRAFT",
                                WEATHER_DELAY == 1 ~ "WEATHER",
                                TRUE ~ "No Delay"))
```

```{r TYPE_DELAY}
pTYPE <- ggplot(data = flights, aes(x = TYPE_DELAY)) +
  geom_bar(fill = "#002D72", color = "#E81828", alpha = 0.8) +
  labs(x = "Type of Delay",
       y = "Frequency",
       title = "Histogram of TYPE_DELAY") + 
  theme(plot.title = element_text(size = 12,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))

pTYPE
```

## Test and Training Set Split

We will use a 80-20 split of training and test sets.

```{r data-split}
set.seed(1234)

flights <- flights %>%
  dplyr::mutate(id = row_number())

train <- flights %>% 
  sample_frac(0.8)
test <- anti_join(flights, train, by = "id")
```

\newpage

# Modeling

## (1) Multiple Linear Regression

### (a) Baesline Linear Model (with AIC Selection)

```{r baseline-lm}
# create linear model with all revelant variables
full_model <- lm(ARR_DELAY ~ 
                    DEP_DELAY + 
                    DAY_OF_WEEK + 
                    OP_CARRIER + 
                    DEST + 
                    CRS_DEP_TIME + 
                    CRS_ARR_TIME + 
                    TAXI_OUT + 
                    TAXI_IN + 
                    TYPE_DELAY, train)

# summary(full_model)
```

```{r, baseline-stepl, message=FALSE}
# use AIC model selection on full model
step_model <- stepAIC(full_model, trace = FALSE)
#summary(step_model)

# update full model
plain_linear_model <- step_model

# plot full model
par(mfrow = c(2,2))
plot(plain_linear_model)
```


### (b) Full Log-Transformed Model 

```{r log-lm}
# fitting models and running ANOVA tests to identify interactions

## first model
lm.01 <- lm(ARR_DELAY ~ DEP_DELAY + DAY_OF_WEEK + OP_CARRIER + DEST + CRS_DEP_TIME + CRS_ARR_TIME + log_TAXI_OUT + log_TAXI_IN + TYPE_DELAY, train)
#plot(lm.01)
#summary(lm.01)

## second model
step_model <- stepAIC(lm.01, direction = "backward", trace = FALSE)
#summary(step_model)

## third model
lm.02 <- lm(ARR_DELAY ~ DEP_DELAY + OP_CARRIER + DEST + CRS_DEP_TIME  + log_TAXI_OUT + log_TAXI_IN + TYPE_DELAY + OP_CARRIER:DEST, train)
#summary(lm.02)
#anova(step_model, lm.02)

## fourth model
lm.03 <- lm(ARR_DELAY ~ DEP_DELAY + OP_CARRIER + DEST + CRS_DEP_TIME  + log_TAXI_OUT + log_TAXI_IN + TYPE_DELAY + OP_CARRIER:DEST + DEST:log_TAXI_IN, train)
#anova(lm.02, lm.03)

## final log model
log_linear_model <- lm(ARR_DELAY ~ DEP_DELAY + OP_CARRIER + DEST + CRS_DEP_TIME  + log_TAXI_OUT + log_TAXI_IN + TYPE_DELAY + OP_CARRIER:DEST + DEST:log_TAXI_IN+log_TAXI_OUT:DEP_DELAY, train)

anova(lm.03, log_linear_model)
summary(log_linear_model)
plot(log_linear_model)
```

## (c) Box-Cox-Transformed Response (w/o Log-Transformed Predictors)

```{r bc-transform}
# adjust response with vertical shift to account for negative responses
## min(train$ARR_DELAY) = -74 --> shift up by constant 75

## transform repsonse
train$adj_ARR_DELAY = train$ARR_DELAY + 75

## fit lm for future Box-Cox use
adj_linear_model <- lm(adj_ARR_DELAY ~ 
                    DEP_DELAY + 
                     OP_CARRIER + 
                     DEST + 
                     CRS_DEP_TIME + 
                     CRS_ARR_TIME + 
                     TAXI_OUT + 
                     TAXI_IN + 
                     TYPE_DELAY,
                   data = train)
```

### Box-Cox Setup
```{r box-cox-notranslm}
# box cox transform (same as case study)
## call EnvStats library
bc_model <- EnvStats::boxcox(adj_linear_model, optimize = TRUE)
# find optimal lambda parameter
bc_lambda <- bc_model$lambda
bc_lambda
#plot(bc_model)

```

```{r add-bc-resp}
# add Box-Cox response as variable to train set
train <- train %>%
  mutate(bc_adj_ARR_DELAY = ((adj_ARR_DELAY^bc_lambda) - 1)/bc_lambda)

# plot training ARR_DELAY
ptrain_ARRDELAY <- ggplot(data = train, aes(x = ARR_DELAY)) +
  geom_histogram(binwidth = 20, fill = "#002D72", color = "#E81828", alpha = 0.8) +
  labs(x = "Arrival Delay",
       y = "Frequency",
       title = "Histogram of ARR_DELAY") + 
  theme(plot.title = element_text(size = 12,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))

# plot Box-Cox training ARR_DELAY
ptrain_bcARRDELAY <- ggplot(data = train, aes(x = bc_adj_ARR_DELAY)) +
  geom_histogram(fill = "#002D72", color = "#E81828", alpha = 0.8) +
  labs(x = "Arrival Delay",
       y = "Frequency",
       title = "Histogram of ARR_DELAY") + 
  theme(plot.title = element_text(size = 12,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))

ptrain_ARRDELAY + ptrain_bcARRDELAY
```
### Final Box-Cox Linear Model

```{r bc-model}
bc_adj_linear_model <- lm(bc_adj_ARR_DELAY ~ 
                    DEP_DELAY + 
                     OP_CARRIER + 
                     DEST + 
                     CRS_DEP_TIME + 
                     CRS_ARR_TIME + 
                     TAXI_OUT + 
                     TAXI_IN + 
                     TYPE_DELAY,
                   data = train)
par(mfrow = c(2,2))
plot(bc_adj_linear_model)
summary(bc_adj_linear_model)

```


## MLR Test Error Metrics & Predictions

LOOKING AT:
- model without any interactions or transformations: ENDED UP PERFORMING THE BEST!!!
- log transformed predictors (taxi_in and taxi_out), interactions, no boxcox transformed response
- box-cox transformed response, no interactions or predicted transformations

```{r lm-predictions}
# predictions
## baseline
test$plain_mlr_pred <- predict(plain_linear_model, test)
## log 
test$log_linear_preds <- predict(log_linear_model, test)
## bc
test$adj_ARR_DELAY = test$ARR_DELAY + 77
test$bc_adj_linear_preds <- predict(bc_adj_linear_model, test)
test <- test %>%
 mutate(adj_linear_preds = ((bc_adj_linear_preds*(bc_lambda) + 1)^(1/bc_lambda)))
test$bc_mlr_pred = test$adj_linear_preds - 77

## histogram of predictions vs. actual
# actual ARR_DELAY in test set
hist(test$ARR_DELAY)
ptest_actual <- ggplot(data = test, aes(x = ARR_DELAY)) +
  geom_histogram(fill = "#E81828", color = "#002D72", alpha = 0.8) +
  labs(x = "Arrival Delay",
       y = "Frequency",
       title = "Histogram of True Predictions of ARR_DELAY") + 
  theme(plot.title = element_text(size = 12,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))

# predicted ARR_DELAY in test set -- baseline lm
ptest_baseline_preds <- ggplot(data = test, aes(x = plain_mlr_pred)) +
  geom_histogram(fill = "#E81828", color = "#002D72", alpha = 0.8) +
  labs(x = "Arrival Delay",
       y = "Frequency",
       title = "Histogram of Predictions of ARR_DELAY from Baseline Linear Model") + 
  theme(plot.title = element_text(size = 12,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))

# predicted ARR_DELAY in test set -- lm with log-trans
hist(test$log_linear_preds)
ptest_lm_log_preds <- ggplot(data = test, aes(x = log_linear_preds)) +
  geom_histogram(fill = "#E81828", color = "#002D72", alpha = 0.8) +
  labs(x = "Arrival Delay",
       y = "Frequency",
       title = "Histogram of Predictions of ARR_DELAY from Linear Model w/ Log-Transformations") + 
  theme(plot.title = element_text(size = 12,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))

# predicted ARR_DELAY in test set -- lm with Box-Cox model
hist(test$log_linear_preds)
ptest_bc_preds <- ggplot(data = test, aes(x = bc_mlr_pred)) +
  geom_histogram(fill = "#E81828", color = "#002D72", alpha = 0.8) +
  labs(x = "Arrival Delay",
       y = "Frequency",
       title = "Histogram of Predictions of ARR_DELAY from Linear Model w/ Box-Cox Response") + 
  theme(plot.title = element_text(size = 12,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))

(ptest_actual + ptest_baseline_preds) / (ptest_lm_log_preds + ptest_bc_preds)


# test MSE calculations
plain_linear_model_MSE <- sum((test$ARR_DELAY - test$plain_mlr_pred)^2, na.rm=T)/length(test$ARR_DELAY)
plain_linear_model_MSE
log_linear_MSE <- sum((test$log_linear_preds-test$ARR_DELAY)^2, na.rm=T)/length(test$ARR_DELAY)
log_linear_MSE
bc_adj_linear_model_MSE <- sum((test$ARR_DELAY - test$bc_mlr_pred)^2, na.rm=T)/length(test$ARR_DELAY)
bc_adj_linear_model_MSE

```


## (2) Generalized Additive Models

### (a) Initial GAM: No Box-Cox on Response
```{r first-gam}
gam00 <- gam(ARR_DELAY ~ DAY_OF_WEEK +
                   OP_CARRIER +
                   s(TAXI_IN) +
                   s(TAXI_OUT) +
                   DEST +
                   s(DEP_DELAY) +
                   s(CRS_DEP_TIME) + 
                   s(CRS_ARR_TIME) + 
                   TYPE_DELAY, data = train)

summary(gam00)
par(mfrow = c(2,3))
plot.gam(gam00, se=TRUE)
```


#### Checking Lineartiy
TAXI_IN may be linear

```{r check-lin-gam}
gam01 <- gam(ARR_DELAY ~ DAY_OF_WEEK +
                   OP_CARRIER +
                   TAXI_IN +
                   s(TAXI_OUT) +
                   DEST +
                   s(DEP_DELAY) +
                   s(CRS_DEP_TIME) + 
                   s(CRS_ARR_TIME) + 
                   TYPE_DELAY, data = train)

anova(gam00, gam01, test = "F")
```

based on anova test, the model with a smoothing spline on TAXI_IN is a better fit

#### More ANOVA Tests

DAY_OF_WEEK, DEST, and CRS_ARR_TIME have very high p-values, so let's try an anova test without including them

```{r}
gam02 <- gam(ARR_DELAY ~ OP_CARRIER +
                   s(TAXI_IN) +
                   s(TAXI_OUT) +
                   s(DEP_DELAY) +
                   s(CRS_DEP_TIME) + 
                   TYPE_DELAY, data = train)

anova(gam00, gam02, test = "F")
```

based on the anova test, the model excluding these variables is a better fit

### Tuned Initial GAM

```{r final-GAM1}
# final fit
summary(gam02)

# diagnostic plots
par(mfrow = c(2,2))
gam.check(gam02)

# predictor plots
par(mfrow = c(2,2))
plot(gam02)
```



## (b) Secondary GAM: Box-Cox on Response

```{r final-bc-gam}
gambc <- gam(bc_adj_ARR_DELAY ~ OP_CARRIER +
                   s(TAXI_IN) +
                   s(TAXI_OUT) +
                   s(DEP_DELAY) +
                   s(CRS_DEP_TIME) + 
                   TYPE_DELAY, data = train)
 
# final BC GAM fit
summary(gambc)

# diagnostic plots
par(mfrow = c(2,2))
gam.check(gambc)

# predictor plots
par(mfrow = c(2,2))
plot(gambc)
```

## GAM Test Error Metrics & Predictions

```{r gam-predictions}
# GAM predictions
## no BC
gam_preds <- predict.gam(gam02, newdata = test)
## BC
gambc_preds <- predict.gam(gambc, newdata = test)
adjgam_preds <- ((gambc_preds*(bc_lambda) + 1)^(1/bc_lambda))
bc_gam_pred = adjgam_preds - 77


## histogram of predictions vs. actual
# actual ARR_DELAY in test set
hist(test$ARR_DELAY)
ptest_actual <- ggplot(data = test, aes(x = ARR_DELAY)) +
  geom_histogram(fill = "#E81828", color = "#002D72", alpha = 0.8) +
  labs(x = "Arrival Delay",
       y = "Frequency",
       title = "Histogram of True Predictions of ARR_DELAY") + 
  theme(plot.title = element_text(size = 12,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))

# predicted ARR_DELAY in test set -- baseline lm
ptest_gam <- ggplot(data = test, aes(x = gam_preds)) +
  geom_histogram(fill = "#E81828", color = "#002D72", alpha = 0.8) +
  labs(x = "Arrival Delay",
       y = "Frequency",
       title = "Histogram of Predictions of ARR_DELAY from GAM") + 
  theme(plot.title = element_text(size = 12,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))

# predicted ARR_DELAY in test set -- lm with log-trans
ptest_gam_bc <- ggplot(data = test, aes(x = bc_gam_pred)) +
  geom_histogram(fill = "#E81828", color = "#002D72", alpha = 0.8) +
  labs(x = "Arrival Delay",
       y = "Frequency",
       title = "Histogram of Predictions of ARR_DELAY from GAM w/ Box-Cox Response") + 
  theme(plot.title = element_text(size = 12,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))

(ptest_actual + ptest_gam) / ptest_gam_bc


# test MSE calculations
gam_MSE <- sum((test$ARR_DELAY - gam_preds)^2, na.rm=T)/length(test$ARR_DELAY)
gam_MSE
gambc_MSE <- sum((test$ARR_DELAY - bc_gam_pred)^2, na.rm=T)/length(test$ARR_DELAY)
gambc_MSE
```

## (3) Tree-Based Models

## (a) Random Forests

By default, ``randomForest()`` uses $p/3$ variables when building a random forest of regression trees.

```{r rf-fitting}
set.seed(1)

# optimal number of predictors (param = mtry) used = 2 based on CV
rf.delay <- randomForest(ARR_DELAY ~ DAY_OF_MONTH +
                   TAXI_IN +
                   TAXI_OUT +
                   DEST +
                   DEP_DELAY +
                   CARRIER_DELAY +
                   NAS_DELAY,
                   data = train, na.action = na.omit, importance = TRUE,
                   ntree=10000, mtry=2)
```


Using the ``importance()`` function, we can view the importance of each variable.

```{r rf-importance}
importance(rf.delay)
```

Two measures of variable importance are reported.  The former is based on the mean decrease in accuracy in predictions on the out of bag samples when a given variable is excluded from the model.  The latter is a measure of the total decrease in node impurity that results from splits over that variable, averaged over all trees (this was plotted in Figure 8.9 in the text).  In the case of regression trees, the node impurity is measured by the training RSS and for classification trees by the deviance.  Plots of these importance measures can be produced using the ``varImpPlot()`` function.

```{r rf-varImp}
varImpPlot(rf.delay)
```

## (b) Boosting 

Here we use the ``gbm()`` package, and within it the ``gbm()`` function, to fit boosted regression trees to the ``train`` data set. We run ``gbm()`` with the option ``distribution = "gaussian"`` since this is a regression problem. The argument ``n.trees = 150`` indicates that we want 150 trees, and the option ``interaction.depth = 3`` limits the depth of each tree.

```{r boost-fititng, message = FALSE, echo=FALSE}

set.seed(1)

# find ideal hyper-parameters through CV
gbmFit <- train(ARR_DELAY ~ DAY_OF_MONTH +
                 TAXI_IN +
                 TAXI_OUT +
                 DEP_DELAY +
                 CARRIER_DELAY +
                 NAS_DELAY +
                 LATE_AIRCRAFT_DELAY, data = train, 
                 method = "gbm")
## note: commented out due to length of output
#gbmFit

# boosted model wiht cross-validated hyper-parameters
boost.delay <- gbm(ARR_DELAY ~ DAY_OF_MONTH +
                 TAXI_IN +
                 TAXI_OUT +
                 DEP_DELAY +
                 CARRIER_DELAY +
                 NAS_DELAY +
                 LATE_AIRCRAFT_DELAY,
                 data = train, distribution = "gaussian",
                 n.trees=150, interaction.depth=3, shrinkage=0.1, cv.folds=10)
```

The ``summary()`` function also provides a relative influence plot and also outputs the relative influence statistics.

```{r boost-summ}
summary(boost.delay)
```

We see that ``DEP_DELAY`` and ``DAY_OF_MONTH`` are by far the most important variables. We can also produce *partial dependence plots* for these two variables.  These plots illustrate the marginal effect of the selected variables on the response after ``integrating`` out the other variables.

```{r boost-imp}
par(mfrow = c(1,2))
plot(boost.delay, i = "DEP_DELAY")
plot(boost.delay, i = "DAY_OF_MONTH")
```



## Trees Test Error Metrics & Predictions
```{r trees-predictions}
# predictions
## random forest
yhat.rf <- predict(rf.delay, newdata = test)
## boosting
yhat.boost <- predict(boost.delay, newdata =test,
                      n.trees = 150)

## histogram of predictions vs. actual
# actual ARR_DELAY in test set
ptest_actual <- ggplot(data = test, aes(x = ARR_DELAY)) +
  geom_histogram(fill = "#E81828", color = "#002D72", alpha = 0.8) +
  labs(x = "Arrival Delay",
       y = "Frequency",
       title = "Histogram of True Predictions of ARR_DELAY") + 
  theme(plot.title = element_text(size = 12,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))

# predicted ARR_DELAY in test set -- baseline lm
ptest_rf <- ggplot(data = test, aes(x = yhat.rf)) +
  geom_histogram(fill = "#E81828", color = "#002D72", alpha = 0.8) +
  labs(x = "Arrival Delay",
       y = "Frequency",
       title = "Histogram of Predictions of ARR_DELAY from Random Forest") + 
  theme(plot.title = element_text(size = 12,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))

# predicted ARR_DELAY in test set -- lm with log-trans
ptest_boost <- ggplot(data = test, aes(x = yhat.boost)) +
  geom_histogram(fill = "#E81828", color = "#002D72", alpha = 0.8) +
  labs(x = "Arrival Delay",
       y = "Frequency",
       title = "Histogram of Predictions of ARR_DELAY from Boosting") + 
  theme(plot.title = element_text(size = 12,hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.x.bottom = element_text(size = 8, face = "italic"),
        axis.title.y.left = element_text(size = 8))

(ptest_actual + ptest_rf) / ptest_boost

# test MSE calculations
rf.MSE <- sum((test$ARR_DELAY - yhat.rf)^2, na.rm=T)/length(test$ARR_DELAY)
rf.MSE
boost_MSE <- sum((test$ARR_DELAY-yhat.boost)^2, na.rm = T)/length(test$ARR_DELAY)
boost_MSE
```

\newpage

## Test Error Tables

```{r full-table}
options(scipen = 5, digits = 4)
model.names <- c("Baseline Linear", "Selected Linear w/ Log-Transformed Predictors", "Selected Linear w/ Box-Cox", "GAM", "GAM w/ Box-Cox", "Random Forest", "Boosting")
model.types <- c("Multiple Linear Regression", "Multiple Linear Regression", "Multiple Linear Regression", "Generalized Additive Model", "Generalized Additive Model", "Tree-Based Regression", "Tree-Based Regression")

#model4.ints <- c("FALSE", "FALSE", "FALSE", "FALSE")

model.mse <- c(plain_linear_model_MSE, log_linear_MSE,
                    bc_adj_linear_model_MSE, gam_MSE,
                    gambc_MSE, rf.MSE, boost_MSE)
model.mse.char <- c("322.46", "333.90", "334.92", "312.30", "317.45", "155.01", "129.80")

#model4.cvmse_var <- c(mlr4_1_cv, mlr4_4_bc_cv, ridge.mom4.cvmse, gam4_bc_gcv)
#model4.cvmse <- c(2.284e+20, "25.66", 25.62, 25.79)


#pctchange4_3 <- round(-((ridge.mom4.cvmse - mlr4_4_bc_cv)/mlr4_4_bc_cv)*100, digits = 4)
#pctchange4_4 <- round(-((gam4_bc_gcv - mlr4_4_bc_cv)/mlr4_4_bc_cv)*100, digits = 4)

#model4.pctchange <- c("---", "---", pctchange4_3, pctchange4_4)

errors.df <- data.frame(model.names, 
                         model.types, 
                         model.mse.char
                         )

 #model4.ints, 
#model4.cvmse, 
                         #model4.pctchange


#colnames(errors.df4) <- c("Model Name", "Model Type", "Interactions?", "Model MSE", "Model CV MSE", "Model Percent Improvement")
errors.df
```

```{r}
library(formattable)
formattable(errors.df,
            col.names = c("Model Name", "Model Type", "Model MSE"),
            list(
  model.names = formatter("span", style = x ~ ifelse(x == "Boosting", 
    style(color = "purple", font.weight = "bold"), NA)),
  model.mse.char = formatter("span", style = x ~ ifelse(x == "129.80", 
    style(color = "purple", font.weight = "bold"), NA)),
  model.types = formatter("span", style = x ~ ifelse(x == "Boosting", 
    style(color = "purple", font.weight = "bold"), NA))
))
```



